---
title: "English Monarchs and Marriages"
author: "Mark Deming"
date: "8/20/2024"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Packages
pacman::p_load(here,        # For here()
               tidyverse,   # For ggplot and dplyr
               plotly,      # For interactive plots
               htmlwidgets, # For saving interactive plots
               MetBrewer)   # For nice color palettes

# Data
tuesdata <- tidytuesdayR::tt_load('2024-08-20')
monarchs_dat <- tuesdata$english_monarchs_marriages_df

```

# Summary
I did not find the data for this week to be very interesting. I felt that there was not much one could do with the data. Accordingly, I looked into the possibility of reading in additional data on, say, length of monarch reign. But I discovered that other data were limited in their coverage of early monarchies.

I ended up creating an interactive version of this plot at x.com: https://x.com/GregoryVdvinne/status/1825743569203679725/photo/1.

Though I am not very excited about the data or the final plot, I did learn some new skills. For instance, I learned some new functions for cleaning datasets (e.g., `across()`, `everything()`, etc.). I also became more familiar with how to generate informative tooltips using the `text` aesthetic. Ultimately, I am most excited about the tooltips: rather than create simple labels, I created complete sentences using `paste()`.

```{r}

# Cleaning
cleaned_monarchs_dat <- monarchs_dat %>% 
  mutate(
         across(everything(), ~na_if(., "-")),
         across(everything(), ~na_if(., "NA")),
         across(everything(), ~na_if(., "?")),
         across(everything(), ~na_if(., "–")),
         across(where(is.character), str_trim),
         across(c(king_age, consort_age, year_of_marriage), ~str_replace_all(., "\\(\\?\\)", "")),
         across(c(king_age, consort_age, year_of_marriage), as.numeric),
         pair_name = paste(king_name, "and", consort_name)
     )

# The cleaning syntax above draws from https://github.com/poncest/tidytuesday/blob/main/2024/Week_34/2024_34.R

```

```{r}

# Color palette
col_palette <- MetBrewer::met.brewer("Juarez")[c(6,1)]

# Plot
p <- 
ggplot(data = subset(cleaned_monarchs_dat, !is.na(cleaned_monarchs_dat$king_age) & !is.na(cleaned_monarchs_dat$consort_age))) +
  geom_segment(aes(x = year_of_marriage, xend = year_of_marriage, y = king_age, yend = consort_age), color = "gray50", size = .25) +
  geom_point(aes(x = year_of_marriage, 
                 y = king_age, 
                 text = paste(pair_name," married in ", year_of_marriage, ".",
                              "\n", king_name, " was ", king_age, ".",
                              "\n", consort_name, " was ", consort_age, ".", sep = ""),
                 color = "Monarch's Age")) +
  geom_point(aes(x = year_of_marriage, 
                 y = consort_age,
                 text = paste(pair_name," married in ", year_of_marriage, ".",
                              "\n", king_name, " was ", king_age, ".",
                              "\n", consort_name, " was ", consort_age, ".", sep = ""), 
                 color = "Consort's Age")) +
  scale_x_continuous(breaks = seq(850,2000,50),
                     labels = seq(850,2000,50)) +
  scale_y_continuous(breaks = seq(0,65,5),
                     labels = seq(0,65,5)) +
  scale_color_manual(name = "Legend\n",
                     values = col_palette) +
  labs(title = "English Monarchs and their Consorts",
       x = "\nYear of Marriage",
       y = "Monarch & Consort Age\n") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        axis.ticks = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 0),
        plot.title.position = "plot")

```

```{r}

saveWidget(ggplotly(p, tooltip = "text"), 
           file = here("2024-08-20", "english_monarchs_interactive_plot.html"))

```
