---
title: "Tidy Tuesday"
subtitle: "Olympic Athletes & Medals"
author: "Mark Deming"
date: "2024-08-06"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load data
tuesdata <- tidytuesdayR::tt_load('2024-08-06')
dat <- tuesdata$olympics

# Load packages
pacman::p_load(tidyverse, 
               extrafont, 
               ggtext,      # for geom_textbox()
               patchwork,   # for arranging plots
               ggpubr)      # for 

```

## Data wrangling

```{r}

# Get female participation for summer Olympics over time
gndr <- dat %>%
  mutate(male = ifelse(sex == "M", 1, 0),
         female = ifelse(sex == "F", 1, 0)) %>%
  filter(season == "Summer") %>%
  group_by(year) %>%
  summarize(male = sum(male),
            female = sum(female)) %>%
  mutate(prop_female = female / (male + female))
  
# Get female participation by sport over time
gndr_sprt <- dat %>%
  mutate(male = ifelse(sex == "M", 1, 0),
         female = ifelse(sex == "F", 1, 0)) %>%
  filter(season == "Summer") %>%
  group_by(year, sport) %>%
  summarize(male = sum(male),
            female = sum(female)) %>%
  mutate(prop_female = female / (male + female))

# Add missing values to archery and tennis
df <- data.frame(year = rep(seq(1896, 2016, 4), times = 2),
                 sport = rep(c("Archery", "Tennis"), each = length(seq(1896, 2016, 4))))

gndr_sprt <- gndr_sprt %>%
  full_join(df, by = c("sport", "year"))

# Add prop_female for year 2000

gndr_sprt <- gndr_sprt %>%
  group_by(sport) %>%
  mutate(prop_female_2000 = ifelse(year == 2000, prop_female, NA)) %>%
  fill(prop_female_2000, .direction = "updown") %>%
  filter(!is.na(prop_female_2000),
         prop_female_2000 != 1,
         sport != "Baseball")

```

## Theming

```{r}

my_theme <- theme(text = element_text(family = "Helvetica Neue Light", size = 7.5),
                  panel.grid.major = element_blank(),
                  panel.grid.minor = element_blank(),
                  axis.ticks = element_blank(),
                  plot.title.position = "plot",
                  axis.text = element_text(angle = 45, hjust = 1, size = 6),
                  strip.background = element_blank(),
                  strip.text = element_text(hjust = 0)
                  )

```

## Plot 1: Female participation over time

```{r}

p1 <-  ggplot(gndr) +
  geom_line(aes(x = year, y = prop_female), color = "gray30", linewidth = .3) +
  scale_x_continuous(limits = c(1896, 2016),
                     breaks = seq(1896, 2016, 4),
                     labels = seq(1896, 2016, 4)) +
  scale_y_continuous(limits = c(0, .50),
                     breaks = seq(0, .50, .05),
                     labels = scales::percent) +
  theme_bw() +
  my_theme +
  labs(title = "Figure 1: Female athletes as a percentage of all summer Olympic athletes",
         x = "\nSummer Olympics year",
         y = "Female althelete percentage\n") 
  
```

# Plot 2: Female participation by sport over time

```{r}

p2 <- ggplot(gndr_sprt) +
  geom_hline(yintercept = .5, color = "gray90", linetype = "dotted") +
  geom_line(aes(x = year, y = prop_female), color = "gray30", linewidth = .3) +
  facet_wrap(~forcats::fct_reorder(sport, -prop_female_2000)) +
  scale_x_continuous(limits = c(1896, 2016),
                     breaks = seq(1896, 2016, 12),
                     labels = seq(1896, 2016, 12)) +
  scale_y_continuous(limits = c(0,.55),
                     breaks = seq(0,.50,.10),
                     labels = scales::percent) +
  theme_bw() +
  my_theme +
  labs(title = "Figure 2: Female athletes as a percentage of all summer Olympic athletes by sport",
         x = "\nSummer Olympics year",
         y = "Female althelete percentage\n",
         caption = "Source: TidyTuesday")

```

## Plot 3: Textbox

```{r}

my_text <- paste("Paris 2024 are the first Olympics to achieve gender parity. Of the 10,500 athletes competing, 50% are men, and 50% are female. As Figure 1 shows, this is the culmination of a slow but relatively steady trend toward greater gender parity over time.\n\n\nOf course, as Figure 2 shows, representation of female athletes has varied across Olympic sports. Organizers made a concerted push toward greater gender parity beginning in 2000.")

text_box <- ggplot(data = tibble(x = 0, y = 1, label = my_text)) +
  aes(x = x, y = y, label = label) +
  geom_textbox(box.color = NA,
               fill = NA,
               width = unit(7.5, "cm"),
               hjust = 0,
               vjust = 1,
               family = "Helvetica Neue Light",
               size = 3.5) +
  scale_x_continuous(limits = c(0, 1), 
                     expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 1), 
                     expand = c(0, 0)) +
  labs(title = "\n\n") +
  theme_void() +
  theme(plot.margin = margin(0, 0, 0, 0))

```

## Arrange plots

```{r}

top <- patchwork::wrap_elements(p1) + 
  patchwork::wrap_elements(text_box) + 
  patchwork::plot_layout(ncol = 2, width = c(4, 3))

bottom <- patchwork::wrap_elements(p2)

my_graphic <- top / bottom + plot_layout(heights = c(2, 3)) + 
  plot_annotation("Gender Parity at the Summer Olympics, 1896-2016", 
                  theme = theme(plot.title = element_text(family = "Helvetica Neue Light")))

```

## Save

```{r}

ggsave("gender_parity_at_olympics.png",
       my_graphic,
       width = 20,
       height = 25,
       units = "cm")

```



# Summary of my process

My initial ideas were very generic: e.g., medals by country. Then, I thought about focusing on the oldest events, and I examined Wikipedia to determine those events. The earliest events weere running events along the lines of the 100M and 1500M. I noted that the pentathlon was introduced in 703 BC and focused on five events linked to soldiering: sprint, horseriding, javelin, shotput, and wrestling. 

Looking at the Wikipedia for the modern pentathlon, I noted two things. First, the specific events differed from the ancient pentathlon and focused on 19th-century cavalryman skills: fencing, swimming, horseriding, pistol shooting, and cross country running. Second, and more importantly, with the introduction of the women's pentathlon in 2000, the event achieved gender parity 24 years before the games as a whole (in Paris 2024!). This prompted me to focus on the story of gender parity in the Olympic games.

I started with the following graphic of the pentathlon and noted how the sudden onset of gender parity in 2000. This is what prompted me to more deeply examine the event's Wikipedia page (What happened in 2000?). On doing so, I discovered that Paris 2024 were the first games as a whole to achieve gender parity.


Let's look at gender parity for the summer olympics over time...

```{r}

p1 <- 
dat %>%
  filter(season == "Summer") %>%
  mutate(male = ifelse(sex == "M", 1, 0),
         female = ifelse(sex == "F", 1, 0)) %>%
  group_by(year) %>%
  summarize(male = sum(male),
            female = sum(female)) %>%
  mutate(prop_female = female/(male + female)) %>%
  ggplot() +
  geom_line(aes(x = year, y = prop_female), color = "gray40") +
  scale_x_continuous(limits = c(1896,2016),
                     breaks = seq(1896,2016,4),
                     labels = seq(1896,2016,4),
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(0,.55),
                     breaks = seq(0,.55,.05),
                     labels = scales::percent,
                     expand = c(0,0)) +
  theme_bw() +
  labs(title = "Figure 1: Female athletes as a percentage of all Olympic athletes",
         x = "\nOlympics year",
         y = "Female althelete percentage\n") +
  theme(text = element_text(family = "Helvetica Neue Light", size = 7.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        plot.title.position = "plot",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6)
        )

```

Event values are gendered male/female. To examine gender parity by event, I will need to remove data on gender from event values.

The tricky bit is to sort out all-female and mixed events (e.g., balance beam and mixed 100m relay, e.g.). 

```{r}

my_dat <- 
dat %>%
  mutate(mal = ifelse(sex == "M", 1, 0),
         fem = ifelse(sex == "F", 1, 0)) %>%
  mutate(event = gsub("Men's", "", event),
         event = gsub("Women's", "", event)) %>%
  group_by(year, season, event) %>%
  summarize(mal = sum(mal),
            fem = sum(fem)) %>%
  group_by(year, season, event) %>%
  mutate(prop_fem = fem/(fem + mal)) %>%
  filter(!grepl("Mixed", event),  # remove mixed events
        mal != 0,
        season == "Summer") %>%
  group_by(season, event) %>%
  summarize(year = year,
            season = season,
            event = event,
            mal = mal,
            fem = fem,
            prop_fem = prop_fem,
            mean_prop = mean(prop_fem))

high_parity <- my_dat %>%
  filter(year == 2000) %>%
  arrange(-prop_fem)

```

Examining parity for different events, it looks like many events achieved parity in 2000, when many women's events were added.

Let's summarize by sport to see if this intuition is correct...

```{r}

p2 <-
dat %>%
  filter(season == "Summer") %>%
  mutate(male = ifelse(sex == "M", 1, 0),
         female = ifelse(sex == "F", 1, 0)) %>%
  group_by(year, sport) %>%
  summarize(male = sum(male),
            female = sum(female)) %>%
  mutate(prop_female = female/(male + female)) %>%
  group_by(sport) %>%
  mutate(prop_female_2000 = ifelse(year == 2000, prop_female, NA)) %>%
  fill(prop_female_2000, .direction = "updown") %>%
  filter(!is.na(prop_female_2000),
         prop_female_2000 != 1,
         sport != "Baseball") %>%
  ggplot(aes(x = year, y = prop_female)) +
  geom_hline(aes(yintercept = .5), linetype = "dotted", color = "gray") +
  geom_line(color = "gray40") +
  facet_wrap(~forcats::fct_reorder(sport, -prop_female_2000)) +
  scale_x_continuous(limits = c(1896,2016),
                     breaks = seq(1896,2016,12),
                     labels = seq(1896,2016,12)) +
  scale_y_continuous(limits = c(0,.55),
                     breaks = seq(0,.55,.10),
                     labels = scales::percent) +
  theme_bw() +
  labs(title = "Figure 2: Female athletes as a percentage of all Olympic athletes by sport",
         x = "\nOlympics year",
         y = "Female althelete percentage\n",
         caption = "Source: TidyTuesday") +
  theme(text = element_text(family = "Helvetica Neue Light", size = 7.5),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        axis.ticks = element_blank(),
        plot.title.position = "plot",
        strip.background = element_blank(),
        strip.text = element_text(hjust = 0),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 6)
        )

```


```{r}

# Create textbox
my_text <- paste("Paris 2024 are the first Olympics to achieve gender parity. Of the 10,500 athletes competing, 50% are men, and 50% are female. As Figure 1 shows, this is the culmination of a slow but relatively steady trend toward greater gender parity over time.\n\n\nOf course, as Figure 2 shows, representation of female athletes has varied across Olympic sports. Organizers made a concerted push toward greater gender parity beginning in 2000.")

text_box <- ggplot(data = tibble(x = 0, y = 1, label = my_text)) +
  aes(x = x, y = y, label = label) +
  geom_textbox(box.color = NA,
               fill = NA,
               width = unit(15, "cm"),
               hjust = 0,
               vjust = 1,
               family = "Helvetica Neue Light",
               size = 7.5) +
  scale_x_continuous(limits = c(0, 1), 
                     expand = c(0, 0)) +
  scale_y_continuous(limits = c(0, 1), 
                     expand = c(0, 0)) +
  labs(title = "\n\n") +
  theme_void() +
  theme(
        plot.margin = margin(0, 0, 0, 0))

# Use patchwork syntax to arrange plots and textbox
top <- patchwork::wrap_elements(p1) + patchwork::wrap_elements(text_box) + patchwork::plot_layout(ncol = 2, width = c(4, 3))
bottom <- patchwork::wrap_elements(p2)
graphic <- top / bottom + patchwork::plot_layout(heights = c(2, 3))

# Save as .png
ggsave("gender_parity_at_olympics.png",
       graphic,
       width = 40,
       height = 50,
       units = "cm")

```


### Reflection
Creating the textbox using ggtext and then arranging the two plots and textbox as a single visualization was tricky. The main challenge was to achieve desirable dimensions for the different plots and overall visualization. I find the trial and error of defining relative dimensions inside `plot_layout()` to be tedious. On dimensions, here is some sample code from nrennie at TidyTuesday: https://github.com/nrennie/tidytuesday/blob/main/2021/24-08-2021/24082021.R.

I would also like to add a title to the final visualization: e.g., "Gender parity at the summer Olympic games." I may be able to achieve this using `text_grob()` from the ggpubr package.

In general, I would like to improve my skills when creating publishable data stories that combine plots and narrative in R. This is my first standalone data story that uses only R, and I learned some new skills: mainly, `grepl`, `geom_textbox`, and arranging plots with patchwork. I would like to further develop these skills to create more appealing data story visualizations with color, improved dimensions, images (e.g., Olympic rings and/or athletes), etc.

Finally, I used piping syntax to summarize and plot the data in a single chunk of code. I find this less useful than summarizing and plotting the data as discrete steps.


