---
title: "Untitled"
author: "Mark Deming"
date: "2026-02-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, extrafont, ggthemes, socviz, janitor, colorspace, scales)

tuesdata <- tidytuesdayR::tt_load('2026-02-17')
tt <- tuesdata$dataset |> clean_names()

```

This was a tricky challenge. It was very difficult to determine what I wanted to convey. I eventually decided to visualize the growth over time of horticultural products with the highest growth. I looked up simple cumulative growth rates as well as compound cumulative growth rates. These yielded slightly different products as the top performers. I went with simple overall growth because it was more intuitive, even if it wasn't the most precise.

In addition to growth rates, I learned about the `slice_min()` and `slice_max()` functions, as well as `first()` and `last()` inside `mutate()`. These will be useful going forward.

The final plots looks okay, but only product 1-4 are very informative. The remaining two are small enough that they appear like flat lines effectively. It's also worth noting that the products I visualized have missing data for several years, but this is not reflected in the visuals themselves given the behavior of `geom_line()`. 

```{r}

dataset_raw <- tt

### |- filter horticulture crops ----
hort_crops <- c("Wine grapes", "Kiwifruit", "Apples", "Avocados")

hort_data <- dataset_raw |>
  filter(measure %in% hort_crops) |>
  select(
    year = year_ended_june,
    crop = measure,
    hectares = value
  ) |>
  arrange(crop, year) |> 
  mutate(is_focus = crop == "Wine grapes")

### |- calculate growth metrics (robust) ----
hort_growth <- hort_data |>
  group_by(crop) |>
  summarise(
    first_year = min(year),
    last_year = max(year),
    first_value = hectares[year == min(year)] |> first(),
    last_value = hectares[year == max(year)] |> first(),
    peak_value = max(hectares, na.rm = TRUE),
    peak_year = year[which.max(hectares)] |> first(),
    growth_pct = (last_value - first_value) / first_value * 100,
    .groups = "drop"
  )

### |- end-point labels ----
end_labels <- hort_data |>
  group_by(crop) |>
  filter(year == max(year)) |>
  ungroup()

### |- crop-specific label offsets ----
label_offsets <- tibble(
  crop = hort_crops,
  dy   = c(0, 0, -1100, 0) 
)

end_labels2 <- end_labels |>
  left_join(label_offsets, by = "crop") |>
  mutate(dy = replace_na(dy, 0))

### |- compute subtitle stats from data  ----
wine_stats <- hort_growth |>
  filter(crop == "Wine grapes") |>
  mutate(
    last_value_round = round(last_value, -2),
    growth_x         = last_value / first_value
  )

wine_growth_x <- round(wine_stats$growth_x, 1)
wine_last_ha <- comma(wine_stats$last_value_round)
  
p <- ggplot(hort_data, aes(x = year, y = hectares, color = crop)) +

  # Geoms
  geom_line(
    aes(alpha = is_focus),
    linewidth = 0.9
  )
p
```



```{r}



# Grab apples
tt |> 
  filter(measure == "Apples") |> 
  ggplot(aes(year_ended_june, value)) +
  geom_line()

apples |>  filter(product == "Apples") |> 
  ggplot(aes(year, hectares)) +
  geom_line()


# Grab plant products
plants <- tt |>
  mutate(value_label = ifelse(value_label == "Hectared", "Hectares", value_label)) |> 
  filter(value_label == "Hectares") |> 
  select(-value_label,
         -value_unit) |> 
  rename(year = year_ended_june,
         product = measure,
         hectares = value)


# Pull out total farmed area as separate column
plants <- plants |> 
  pivot_wider(id_cols = year,
              names_from = product,
              values_from = hectares) |> 
  pivot_longer(cols = -c(year, `Total Area of Farms`),
               names_to = "product",
               values_to = "hectares") |> 
  rename(total_farmed_area = `Total Area of Farms`)


# Find first, last, max, min values
plants <- plants |> 
  group_by(product) |> 
  mutate(first_year = first(year[!is.na(hectares)]),
         last_year = last(year[!is.na(hectares)]),
         max_value = max(hectares, na.rm = T),
         min_value = min(hectares, na.rm = T)) |> 
  filter(year > 1981)


# Create compound annual growth rates
plants <- plants |> 
  arrange(product, year) |>
  group_by(product) |>
  mutate(
    simple_growth = (hectares[year == last_year] - hectares[year == 1982]) / hectares[year == 1982],
    growth_rate = (hectares / lag(hectares)) - 1,
    growth_rate = replace_na(growth_rate, 0),

    # Handle lag(value) == 0 explicitly (choose a rule)
    growth_rate = if_else(is.infinite(growth_rate), NA_real_, growth_rate),

    growth_multiplier = 1 + growth_rate,
    cum_growth_multiplier = cumprod(replace_na(growth_multiplier, 1)),
    final_cum_rate = last(cum_growth_multiplier) - 1
  ) |>
  ungroup()


# Grab high and low growth agricultural products
growth_tbl <- plants |> 
  arrange(product, year) |> 
  ungroup() |> 
  group_by(product) |> 
  summarize(#final_cum_rate = last(final_cum_rate),
            simple_growth = last(simple_growth),
            .groups = "drop")

#top_bottom_products <- bind_rows(
#  growth_tbl |> slice_max(simple_growth, n = 15),
#  growth_tbl |> slice_min(simple_growth, n = 15)
#)

# Top only
top_products <- 
  growth_tbl |> slice_max(simple_growth, n = 30)

drops <- c("Maize (area sown)", "Hectares effluent sprayed over", "Flowers and Foliage (Outdoor)",
           "Other Outdoor Horticultural Crops", "New Area Planted in Production Forest", 
           "Other land", "Wheat (area sown)", "Barley (area sown)", "Plantations of Exotic Trees",
           "Tussock or Danthonia", "Crops, Grain, Nursery, Vegetables and Fruit", "Oats (area sown)",
           "Grassland and Lucerne", "Grain", "Field peas (area sown)", 
           "Crops, Grain, Nursery, Vegetables and Fruit", "	Plantations of Exotic Trees")

#top_bottom_products <- top_bottom_products |> 
#  filter(product %nin% drops)

#top_bottom_products <- bind_rows(
#  top_bottom_products |>  slice_max(simple_growth, n = 6),
#  top_bottom_products |>  slice_min(simple_growth, n = 6)
#)

top_products <- top_products |> 
  filter(product %nin% drops) |> 
  slice_max(simple_growth, n = 6)

# Add years back in 
df_top <- plants %>%
  semi_join(top_products, by = "product")


# Remove NA values(?)
df_top <- df_top |> 
  filter(!is.na(hectares))


# ID Winners & Losers
#winners <- c("Wine grapes", "Kiwifruit", "Apples", "Avocados", "Asparagus", "Cherries")
#losers <- c("Peaches", "Blueberries", "Tangelos", "Tamarillos", "Raspberries", "Grapefruit/Goldfruit")


# Add Ha value for 2024
df_top <- df_top |> 
  group_by(product) |> 
  mutate(last_value = last(hectares),
         first_value = first(hectares))

# Color palette
pal <- MetBrewer::met.brewer("Lakota")
swatchplot(pal)


# Plot 1 - WINNERS
p <-
ggplot(df_top,
       aes(year, hectares, group = product)) +
  
  geom_line(aes(color = product), size = .75) +
  
  facet_wrap(~fct_reorder(product, -max_value), ncol = 2) +
  
  scale_x_discrete(breaks = seq(1982, 2024, 10),
                   expand = expansion(mult = c(.2, .2))) +
  
  scale_y_continuous(limits = c(0, 40000),
                     breaks = seq(0, 40000, 10000),
                     expand = expansion(mult = c(.1, .1)),
                     labels = scales::label_number(scale_cut = scales::cut_short_scale())) +
  
  scale_color_manual(values = pal, 
                     guide = F) +
  
  geom_text(aes(x = 2024, y = last_value, 
                label = paste(scales::comma(last_value), "Ha.")),
            size = 2.75,
            family = "Raleway",
            color = "gray40",
            nudge_x = .5,
            hjust = "left") +
  
    geom_text(aes(x = 1982, y = first_value, 
                label = paste(scales::comma(first_value), "Ha.")),
            size = 2.75,
            family = "Raleway",
            color = "gray40",
            nudge_x = -.5,
            hjust = "right") +
  
  labs(title = "What's growing in New Zealand?",
       subtitle = "Planted area in Hectares for horticultural crops with highest growth, 1982-2024\n",
       x = "", y = "") +
  
  theme_minimal() +
  
  theme(text = element_text(family = "Raleway"),
        panel.grid.minor = element_blank(),
        panel.grid.major.y = element_line(size = .25),
        panel.grid.major.x = element_blank(),
        strip.text = element_text(hjust = 0),
        plot.title.position = "plot")

p

ggsave("top_growing_agr_product_line_graph.png",
       p,
       width = 8,
       height = 6,
       dpi = 300
       )

```
